Neural Connectome of the Ctenophore Statocyst

### 2025-05-16 manuscript/Manuscript.qmd
MS: fix ref fields

---
title: "Neural Connectome of the Ctenophore Statocyst"
subtitle: |
  **Kei Jokura**$^{1,2,3,4,5*}$, **Sanja Jasek**$^{1,2,6}$, **Lara Kewalow**$^{6}$, **Pawel Burkhardt**$^{7}$, **Gáspár Jékely**$^{1,2,6,*}$  
  <br><br>  
  $^{1}$ Living Systems Institute, University of Exeter, Exeter, EX4 4QD, United Kingdom  
  <br>
  $^{2}$ Biosciences, Faculty of Health and Life Sciences, University of Exeter, Exeter EX4 4QD, UK
  <br>  
  $^{3}$ Grass Laboratory, Marine Biological Laboratory, Woods Hole, MA 02543, USA  
  <br> 
  $^{4}$ Exploratory Research Center on Life and Living Systems (ExCELLS), Okazaki, 444-8787, Japan  
  <br>  
  $^{5}$ National Institute for Basic Biology (NIBB), Okazaki, 444-8585, Japan  
  <br> 
  $^{6}$ Heidelberg University, Centre for Organismal Studies (COS), 69120 Heidelberg, Germany  
  <br>
  $^{7}$ Michael Sars Centre, University of Bergen, Norway
  <br><br>  
  **$^{*}$ Correspondence:**  
  [jokura@nibb.ac.jp](mailto:jokura@nibb.ac.jp),  
  [gaspar.jekely@cos.uni-heidelberg.de](mailto:gaspar.jekely@cos.uni-heidelberg.de)
mainfont: Arial
monofont: DejaVu Sans Mono #Times New Roman
fontsize: 11pt
linestretch: 1.2
linkcolor: blue
documentclass: article
max-width: 100%
format:
  html:
    embed-resources: true 
  pdf:
    pdf-engine: xelatex
    number-sections: false
    margin-left: 0.8in
    margin-right: 0.8in
    margin-top: 0.8in
    margin-bottom: 0.8in
    toc: false
    keep-tex: true
    include-in-header:
      text: |
        \usepackage{lineno}
    include-before-body:
      text: |
        \linenumbers
bibliography: references.bib #add bibliography in this file
csl: elife.csl #change according to journal
---

@@

To address the lack of structural understanding of how ctenophores control behavior, we focused on a unique sensory structure—the aboral organ. This gravity-sensing organ contains a statocyst supported by four mechanosensory balancer cells, which act as pacemakers for comb plate beating and posture control [@HORRIDGE1965; @Tamm1980]. Ultrastructural and immunohistochemical studies have revealed that the aboral organ contains diverse sensory and neural components, including a distinct neural structure termed the deep nerve net [@aronova1974electron; @Hernandez-Nicaise1968; @Jager2011]. A key feature of balancer cilia is their bidirectional, mechanoresponsive control of beat frequency. Depending on the direction of deflection and the animal’s geotactic state, balancer cilia may either increase or decrease their activity, indicating a dynamic response system not fixed like vertebrate hair cells [@HORRIDGE1965; @Tamm1982; @Tamm_2014b]. Previous studies have implicated membrane potential, intracellular calcium concentration, and synaptic input in modulating this response [@Lowe1997]. In particular, deflection-induced excitation can be abolished by calcium channel inhibitors or calcium-free conditions [@Lowe1997], while high extracellular K⁺ depolarizes and activates cilia in the absence of mechanical stimuli [@Lowe1997]. Local calcium application further suggests that Ca²⁺ influx at the ciliary base may be required for the excitatory response [@Lowe1997]. These findings support the hypothesis that neural modulation plays a critical role in shaping balancer cell output [@Lowe1997].

@@

Classical neural staining techniques did not provide clear images of neurons at the aboral organ. However, ultrastructural studies revealed morphological evidence for neural elements, including synapses that are located on the epithelial floor of the aboral organ—a region where epithelial cells are densely packed and form a floor-like structure [@Horridge_1964; @Hernandez_Nicaise_1973]. In our dataset, we identified synapses based on the previously described pre-synaptic triad morphology consisting of a single layer of vesicles, a cisterna of smooth endoplasmic reticulum and an associated mitochondrion. At synaptic sites, we marked mitochondria as synaptic nodes (orange in CATMAID) and connected this node to the nearest node in postsynaptic cells across the regions where synaptic vesicles aligned with the presynaptic cell's membrane (light blue arrows in CATMAID) (Supplementary Figure S2A). We could not detect specialized post-synaptic structures, in agreement with previous studies [@Hernandez_Nicaise_1973]. Synapses were identified as either monoadic or polyadic, with one pre-synaptic neuron forming connections with one or multiple post-synaptic cells (Supplementary Figure S2B).

@@

To investigate the regulation of balancer function, we carried out a high-speed video-microscope analysis of balancer cilia. Previous studies [@Tamm1980; @Tamm1982; @Lowe1997] have established that balancer cilia function as mechanoreceptors, with their beating frequency modulated by inclination. Tamm also suggested that differences in statolith morphology and the shape of balancer cilia between the tentacular and sagittal planes could lead to different forces exerted on cilia by the statolith [@Tamm_2014; @Tamm2015]. To further explore this, we used a tilted microscope with a vertical stage where we mounted immobilised cydippid larvae with their aboral-oral axis aligned in different orientations relative to the gravity vector. We then compared larvae oriented in different angles and either with their sagittal or tentacular plane parallel to the sample stage.

@@

In our volumetric reconstruction, bridge cells were identified as the only cell type providing synaptic input to the ANN network, positioning them as key upstream regulators of the ciliary coordination system in the aboral organ. Originally described morphologically by Tamm et al. (2002), bridge cells extend thin projections toward the base of the balancer cilia [@Tamm_2002; @Tamm_2014b]. Based on this spatial relationship, they may receive information about the physical state of the cilia—such as mechanical load or phase—or possibly local field potential (LFP) changes [@sheu2022; @jurisch-yaksi2024]. These features suggest that bridge cells may act as transducers, converting local ciliary signals into neural inputs directed toward the ANN circuit.

>>>>

---
title: "Neural Connectome of the Ctenophore Statocyst"
subtitle: |
  **Kei Jokura**$^{1,2,3,4,5*}$, **Sanja Jasek**$^{1,2,6}$, **Lara Kewalow**$^{6}$, **Pawel Burkhardt**$^{7}$, **Gáspár Jékely**$^{1,2,6,*}$  
  <br><br>  
  $^{1}$ Living Systems Institute, University of Exeter, Exeter, EX4 4QD, United Kingdom  
  <br>
  $^{2}$ Biosciences, Faculty of Health and Life Sciences, University of Exeter, Exeter EX4 4QD, UK
  <br>  
  $^{3}$ Grass Laboratory, Marine Biological Laboratory, Woods Hole, MA 02543, USA  
  <br> 
  $^{4}$ Exploratory Research Center on Life and Living Systems (ExCELLS), Okazaki, 444-8787, Japan  
  <br>  
  $^{5}$ National Institute for Basic Biology (NIBB), Okazaki, 444-8585, Japan  
  <br> 
  $^{6}$ Heidelberg University, Centre for Organismal Studies (COS), 69120 Heidelberg, Germany  
  <br>
  $^{7}$ Michael Sars Centre, University of Bergen, Norway
  <br><br>  
    **$^{*}$ Correspondence:**  ![](pictures/emails.png){width=30em}
mainfont: Arial
monofont: DejaVu Sans Mono #Times New Roman
fontsize: 11pt
linestretch: 1.2
linkcolor: blue
documentclass: article
max-width: 100%
format:
  html:
    embed-resources: true 
  pdf:
    pdf-engine: xelatex
    number-sections: false
    margin-left: 0.8in
    margin-right: 0.8in
    margin-top: 0.8in
    margin-bottom: 0.8in
    toc: false
    keep-tex: true
    include-in-header:
      text: |
        \usepackage{lineno}
    include-before-body:
      text: |
        \linenumbers
bibliography: references.bib #add bibliography in this file
csl: elife.csl #change according to journal
---

@@

To address the lack of structural understanding of how ctenophores control behavior, we focused on a unique sensory structure—the aboral organ. This gravity-sensing organ contains a statocyst supported by four mechanosensory balancer cells, which act as pacemakers for comb plate beating and posture control [@HORRIDGE1965; @Tamm1980]. Ultrastructural and immunohistochemical studies have revealed that the aboral organ contains diverse sensory and neural components, including a distinct neural structure termed the deep nerve net [@aronova1974electron; @Hernandez_Nicaise1968; @Jager2011]. A key feature of balancer cilia is their bidirectional, mechanoresponsive control of beat frequency. Depending on the direction of deflection and the animal’s geotactic state, balancer cilia may either increase or decrease their activity, indicating a dynamic response system not fixed like vertebrate hair cells [@HORRIDGE1965; @Tamm1982; @tamm2014b]. Previous studies have implicated membrane potential, intracellular calcium concentration, and synaptic input in modulating this response [@Lowe1997]. In particular, deflection-induced excitation can be abolished by calcium channel inhibitors or calcium-free conditions [@Lowe1997], while high extracellular K⁺ depolarizes and activates cilia in the absence of mechanical stimuli [@Lowe1997]. Local calcium application further suggests that Ca²⁺ influx at the ciliary base may be required for the excitatory response [@Lowe1997]. These findings support the hypothesis that neural modulation plays a critical role in shaping balancer cell output [@Lowe1997].

@@

Classical neural staining techniques did not provide clear images of neurons at the aboral organ. However, ultrastructural studies revealed morphological evidence for neural elements, including synapses that are located on the epithelial floor of the aboral organ—a region where epithelial cells are densely packed and form a floor-like structure [@Horridge_1964; @Hernandez_Nicaise1973a]. In our dataset, we identified synapses based on the previously described pre-synaptic triad morphology consisting of a single layer of vesicles, a cisterna of smooth endoplasmic reticulum and an associated mitochondrion. At synaptic sites, we marked mitochondria as synaptic nodes (orange in CATMAID) and connected this node to the nearest node in postsynaptic cells across the regions where synaptic vesicles aligned with the presynaptic cell's membrane (light blue arrows in CATMAID) (Supplementary Figure S2A). We could not detect specialized post-synaptic structures, in agreement with previous studies [@Hernandez_Nicaise1973a]. Synapses were identified as either monoadic or polyadic, with one pre-synaptic neuron forming connections with one or multiple post-synaptic cells (Supplementary Figure S2B).

@@

To investigate the regulation of balancer function, we carried out a high-speed video-microscope analysis of balancer cilia. Previous studies [@Tamm1980; @Tamm1982; @Lowe1997] have established that balancer cilia function as mechanoreceptors, with their beating frequency modulated by inclination. Tamm also suggested that differences in statolith morphology and the shape of balancer cilia between the tentacular and sagittal planes could lead to different forces exerted on cilia by the statolith [@Tamm_2014c; @tamm2015]. To further explore this, we used a tilted microscope with a vertical stage where we mounted immobilised cydippid larvae with their aboral-oral axis aligned in different orientations relative to the gravity vector. We then compared larvae oriented in different angles and either with their sagittal or tentacular plane parallel to the sample stage.

@@

In our volumetric reconstruction, bridge cells were identified as the only cell type providing synaptic input to the ANN network, positioning them as key upstream regulators of the ciliary coordination system in the aboral organ. Originally described morphologically by Tamm et al. (2002), bridge cells extend thin projections toward the base of the balancer cilia [@Tamm_2002; @tamm2014b]. Based on this spatial relationship, they may receive information about the physical state of the cilia—such as mechanical load or phase—or possibly local field potential (LFP) changes [@sheu2022; @jurisch-yaksi2024]. These features suggest that bridge cells may act as transducers, converting local ciliary signals into neural inputs directed toward the ANN circuit.
